---
title: "math 189 Project 1"
output: html_notebook
---
# Case Study 1
Caitlyn Bryant - Applied Math, 4th yr
Edward Chao - Math: Statistics, 4th yr
Kieran Mann - Math/Econ, 4th yr 
Matthew Kucirek - Applied Math, 4th yr

# Setup
Set working directory and read in babies file

```{r}

setwd("~/RStudio") # desired directory as argument
bigdata <- read.table("./babies23.txt", header=TRUE)

summary(bigdata)


```


# Assumption: Discard observations that aren't valid 
9 is recorded for no observation on smoking, 999 for no observation on weight, 98 for no observation on number smoked

```{r}

#filter out useless rows using subset
bigdata = subset(bigdata, wt != 999 & smoke != 9 & number != 98)

```


# Investigation

 - Summarize numerically the two distributions of birth weight for babies born to
women who smoked during their pregnancy and for babies born to women who
did not smoke during their pregnancy.
 - Use graphical methods to compare the two distributions of birth weight.
 - Asses the importance of the difference you found in your three types of
comparisons (numerical, graphical, incidence). Summarize your findings and
relate them to other studies.

# Meaning of the "smoke" variable

0 - no 
1 - yes
2 - up until preg
3 - once, but not while I was pregnant
9 - unknown (ommitted)

# Graphical Analysis

```{r}

install.packages("dplyr") #used to handle data tables
library(dplyr)

#HISTOGRAM
#creating data tables to be made into histograms
nonsmokers = filter(bigdata, smoke == 0)
smokers = filter(bigdata, smoke == 1 | smoke == 2 | smoke == 3)
now = filter(smokers, smoke == 1)
preg = filter(smokers, smoke == 2)
once = filter(smokers, smoke == 3)

#histograms of smokers/nonsmokers 
nosmoke = nonsmokers[,7]
smoke = smokers[,7]
hnosmoke = hist(nosmoke, main="Histogram of Baby Weights to Smokers", xlab="Baby Weight in Ounces")
hsmoke = hist(smoke, main="Histogram of Baby Weights to Non-Smokers", xlab="Baby Weight in Ounces")

#histograms of frequency smoked
bwtnow = now[,7]
bwtpreg = preg[,7]
bwtonce = once[,7]
hbwtnow = hist(bwtnow, main="", xlab = "Smoked Before and Throughout Pregnancy", cex.lab=1.5, cex.axis=1.5, cex.sub=1.5)
hbwtpreg = hist(bwtpreg, main="", xlab = "Smoked Up Until Pregnancy", cex.lab=1.5, cex.axis=1.5, cex.sub=1.5)
hbwtonce = hist(bwtonce, main="", xlab = "Smoked Once", cex.lab=1.5, cex.axis=1.5, cex.sub=1.5)

#Box Plots
bwtsmokeplot = boxplot(wt~smoke, bigdata)
bnumber = filter(bigdata, number != 98)
bwtnumberplot = boxplot(wt~number, bnumber, main="Birthweight vs. Number of Cigarettes", xlab="Number of Cigarettes Smoked a Day", ylab="Infant Birth Weight")

#QQ Plots
#mothers BMI vs. bwt
install.packages("measurements")
library(measurements)
measures = filter(bigdata, wt.1 != 999, ht != 99)
bmi.mom = mutate(measures, BMI = (conv_unit(measures[,13], "lbs", "kg")/(conv_unit(measures[,12], "inch", "m"))^2))
plot(bmi.mom$BMI, bmi.mom$wt)
qqplot(bmi.mom$BMI, bmi.mom$wt, main="Mother's BMI vs Infant Birth Weight", xlab="Mother's BMI", ylab="Infant Birth Weight", xlim = c(0, 180), ylim = c(0,180))
abline(c(0,1))

#dad's BMI vs. bwt
dmeasures = filter(bigdata, dwt != 999, dht != 99)
bmi.dad = mutate(dmeasures, BMI = (conv_unit(dmeasures[,18], "lbs", "kg")/(conv_unit(dmeasures[,17], "inch", "m"))^2))
plot(bmi.dad$BMI, bmi.dad$wt)
qqplot(bmi.dad$BMI, bmi.dad$wt, xlim = c(0, 180), ylim = c(0,180))
abline(c(0,1))

```


# Frequency analysis 

We created a rudimentary frequency analysis function to use in this section

```{r}

# define the function used here for frequency analysis
analyzeFreq <- function (vector, resolution, range) {
  toReturn <- vector
  # find the total range of observations
  if (!range) totalrange <- range(vector)
  else totalrange = range
  
  # break that range into intervals by the given reosultion
  breaks = seq(totalrange[1], totalrange[2], by=resolution)
  
  # cut and create a frequency table
  toReturn.cut = cut(vector, breaks, right=FALSE)
  toReturn.frequency = table(toReturn.cut)

  
  return (toReturn.frequency)
}

```

# Create Frequency Analysis Tables/Histograms

```{r}
# first, overall frequency distribution for all babies
# range of values for birthweight

# call the function oreviously defined up to get a frequency table

allweights <- bigdata$wt

totalRange <- range(allweights)

allFreq <- analyzeFreq(allweights, 11, totalRange)

View(allFreq)

plot(allFreq, main="Frequency of birthweight by interval (all mothers)", sub="For all mothers, regardless of smoking status", xlab="Birth weight by interval, grams", ylab="Frequency")

smokersweights <- bigdata$wt[bigdata$smoke == 1]

smokers <- analyzeFreq(smokersweights, 11, totalRange)

View(smokers)

plot(smokers)

# group non smokers with those that quit and distant past
nonsmokersweights <- bigdata$wt[bigdata$smoke != 1]

nonsmokers <- analyzeFreq(nonsmokersweights, 11, totalRange)

View(nonsmokers)
plot(nonsmokers)


# now analyze the frequency of cigaretter smoking among mothers 
# adjust the observation to reality
numsmoked <- (bigdata$number * 5)

numsmoked
smokeFreq <- analyzeFreq(numsmoked, 1, c(55, 176))


h = hist(numsmoked)

# make the histogram of density, not total frequency
h$density = h$counts/sum(h$counts)*100
plot(h,freq=FALSE, xlab="Cigarettes smoked, interval", ylab="Percentage of mothers")

# frequency of early babies among smokers

earlySmokers <- bigdata$gestation[bigdata$smoke == 1 & bigdata$gestation != 999]

# make days into weeks
earlySmokers <- earlySmokers / 7

earlySmokers.hist <- hist(earlySmokers)

# make the histogram of density, not total frequency
earlySmokers.hist$density = earlySmokers.hist$counts/sum(earlySmokers.hist$counts)*100

plot(earlySmokers.hist,freq=FALSE, main="Frequency of Gestation Period, Smokers", xlab="Gestation Period", ylab="Percentage of mothers")

# frequency of early babies among non smokers

earlyNonsmokers <- bigdata$gestation[bigdata$smoke != 1 & bigdata$gestation != 999]

# make days into weeks
earlyNonsmokers <- earlyNonsmokers / 7

earlyNonsmokers.hist <- hist(earlyNonsmokers)

# make the histogram of density, not total frequency
earlyNonsmokers.hist$density = earlyNonsmokers.hist$counts/sum(earlyNonsmokers.hist$counts)*100

plot(earlyNonsmokers.hist,freq=FALSE, main="Frequency of Gestation Period, NonSmokers", xlab="Gestation Period", ylab="Percentage of mothers")


# frequency of low birth weight among early babies

allGestationWks <- bigdata$gestation[bigdata$gestation != 999 ] /7

weights <- bigdata$wt[bigdata$gestation != 999]
earlyLight <- weights[allGestationWks < 37]

earlyLight.hist <- hist(earlyLight)

# make the histogram of density, not total frequency
earlyLight.hist$density = earlyLight.hist$counts/sum(earlyLight.hist$counts)*100

plot(earlyLight.hist,freq=FALSE, main="Frequency of Birthweight, Babies before 37 weeks", xlab="Birth weight, grams", ylab="Percentage of mothers")


earlyLight.table <- analyzeFreq(earlyLight, 11, totalRange)

View(earlyLight.table)


```

# Numerical Analysis 

```{r}

#we will need moments to calculate moments 
library("moments")

options(max.print=1000000)
#read in the data 
data <- read.table("babies23.txt",header=1)
#View(data)

#0 - not smoke 
#1 - smoke 
#2 smoke up until pregnancy 
#3 smoked once, not while pregnant 

#separate based on smoking category 

d0 <- data[data$smoke == 0,]
#View(d0)
d1 <- data[data$smoke == 1,]
#View(d1)
d2 <- data[data$smoke == 2,]
#View(d2)
d3 <- data[data$smoke == 3,]
#View(d3)

d9 <- data[data$smoke == 9,]


#caluclate respective means, variance, skew, kurtosis 

m0 <- mean(d0$wt)
sd0 <- sqrt(var(d0$wt))
skew0 <- skewness(d0$wt)
kurt0 <- kurtosis(d0$wt)

s0 <- c(m0,sd0,skew0,kurt0)

m1 <- mean(d1$wt)
sd1 <- sqrt(var(d1$wt))
skew1 <- skewness(d1$wt)
kurt1 <- kurtosis(d1$wt)

s1 <- c(m1,sd1,skew1,kurt1)

m2 <- mean(d2$wt)
sd2 <- sqrt(var(d2$wt))
skew2 <- skewness(d2$wt)
kurt2 <- kurtosis(d2$wt)

s2 <- c(m2,sd2,skew2,kurt2)

m3 <- mean(d3$wt)
sd3 <- sqrt(var(d3$wt))
skew3 <- skewness(d3$wt)
kurt3 <- kurtosis(d3$wt)

s3 <- c(m3,sd3,skew3,kurt3)

m9 <- mean(d9$wt)
sd9 <- sqrt(var(d9$wt))
skew9 <- skewness(d9$wt)
kurt9 <- kurtosis(d9$wt)

s9 <- c(m9,sd9,skew9,kurt9)


#the matrix m is a numerical summary 
#of the distributions of our babys weights 
# separtated by the 4 categories posed in the start 
m <- cbind(s0,s1,s2,s3,s9)

# used to obtain the quartiles/extremes of the data 

summary(d0$wt)
summary(d1$wt)
summary(d2$wt)
summary(d3$wt)

#agostino test runs a hypothesis test on the skewess,
#testing for normality

agostino.test(d0$wt)
agostino.test(d1$wt)
agostino.test(d2$wt)
agostino.test(d3$wt)

#the anscombe test runs a hypothesis test on kurtosis,
#again testing for normailty 

anscombe.test(d0$wt)
anscombe.test(d1$wt)
anscombe.test(d2$wt)
anscombe.test(d3$wt)

#separating underweight and overweight babies 
#while keeping them in their respective 
#smoking category 

bw0 <- d0$wt[d0$wt < 89]
uw0 <- d0$wt[d0$wt > 88]

bw1 <- d1$wt[d1$wt < 89]
uw1 <- d1$wt[d1$wt > 88]

bw2 <- d2$wt[d2$wt < 89]
uw2 <- d2$wt[d2$wt > 88]

bw3 <- d3$wt[d3$wt < 89]
uw3 <- d3$wt[d3$wt > 88]

bw9 <- d9$wt[d9$wt < 89]
uw9 <- d9$wt[d9$wt > 88]

#calculating proportions of babies in each category 
#for example b0 is proportion of babies who are 
#below weight and in smoke category 0 

b0 <- length(bw0)/1236
u0 <- length(uw0)/1236

b1 <- length(bw1)/1236
u1 <- length(uw1)/1236

b2 <- length(bw2)/1236
u2 <- length(uw2)/1236

b3 <- length(bw3)/1236
u3 <- length(uw3)/1236

b9 <- length(bw9)/1236
u9 <- length(uw9)/1236

#creates a matrix of probabilities 

brow <- c(b0,b1,b2,b3,b9)
urow <- c(u0,u1,u2,u3,u9)

mp <- rbind(brow,urow)
View(mp)

#calculating probabilities of the column events happening 
#to later calulcate conditional probabilities 

p0 <- sum(mp[,1])
p1 <- sum(mp[,2])
p2 <- sum(mp[,3])
p3 <- sum(mp[,4])
p9 <- sum(mp[,5])

p <- c(p0,p1,p2,p3,p9)

mp <- rbind(mp,p)

pw <- c(sum(brow),sum(urow),sum(p))

mp <- cbind(mp,pw)

#creates matrix of conditional probabilites 

cp <- c(mp[1,1]/mp[3,1],
        mp[1,2]/mp[3,2],
        mp[1,3]/mp[3,3],
        mp[1,4]/mp[3,4],
        mp[1,5]/mp[3,5])

cp1 <- c(mp[2,1]/mp[3,1],
        mp[2,2]/mp[3,2],
        mp[2,3]/mp[3,3],
        mp[2,4]/mp[3,4],
        mp[2,5]/mp[3,5])

cpt <- rbind(cp,cp1)

#looking at underweight and preterm babies 

d <- data[data$gestation<259 & data$wt <89,]

dzero <-d$smoke[d$smoke == 0]
done <- d$smoke[d$smoke == 1]
dtwo <-d$smoke[d$smoke == 2]
dthree <-d$smoke[d$smoke == 3]

#repeats the process of finding conditional probabilities 

row_uwpt <- c(length(dzero)/1236,length(done)/1236,
              length(dtwo)/1236,length(dthree)/1236)

row_nuwpt <- c((length(d0$smoke)-length(dzero))/1236,
                (length(d1$smoke)-length(done))/1236,
                (length(d2$smoke)-length(dtwo))/1236,
                (length(d3$smoke)-length(dthree))/1236
               )

puwpt <- rbind(row_uwpt,row_nuwpt)
h <- c(sum(puwpt[,1]),sum(puwpt[,2]),sum(puwpt[,3]),sum(puwpt[,4]))
puwpt <- rbind(puwpt,h)
View(puwpt)

prob_p <- c(puwpt[1,1]/puwpt[3,1],puwpt[1,2]/puwpt[3,2],
            puwpt[1,3]/puwpt[3,3],puwpt[1,4]/puwpt[3,4])

prob_p2 <-  c(puwpt[2,1]/puwpt[3,1],puwpt[2,2]/puwpt[3,2],
              puwpt[2,3]/puwpt[3,3],puwpt[2,4]/puwpt[3,4])         
prob_puwpt <- rbind(prob_p,prob_p2)
View(prob_puwpt)

```

